---
title: "clean"
format: html
editor: visual
---

## Loading necessary libraries

```{r}
install.packages("tidyverse")
```

```{r}
library(tidyverse)
library(dplyr)
library(stringr)
library(tidyr)
```

## Initial exploration

```{r}
# Initial inspection
glimpse(TCGA)
```

```{r}
head(TCGA)
summary(TCGA)
colnames(TCGA)
```

```{r}
# Remove first column since it is redundant
TCGA <- TCGA[,-1]
```

## Missing values

```{r}
# Check for missing values and gather the number of missing values per column
missing_values_detailed <- TCGA %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(cols = everything(), names_to = "attribute", values_to = "missing_count") %>%
  arrange(desc(missing_count))

# View detailed missing values summary
print(missing_values_detailed)
```

```{r}
# Visualize the missing values using a simple plot
missing_values_detailed %>%
  filter(missing_count > 0) %>%
  ggplot(aes(x = reorder(attribute, missing_count), y = missing_count)) +
  geom_bar(stat = "identity") +
  coord_flip() + # Flip coordinates for horizontal bar plot
  labs(x = "Attributes", y = "Number of Missing Values", title = "Missing Values per Attribute") +
  theme_minimal()
```

### Special values

In the data values like "\[Not Available\]", "\[Not Applicable\]", and "\[Discrepancy\]" are non-NA values but represent missing information in a practical sense and will be treated like categorical values.

For instance, "\[Not Available\]" could mean that the data was never collected, while "\[Not Applicable\]" could mean the question or data point was not relevant for that particular observation. In such cases, they are not simply 'missing' but convey a distinct meaning that could be important for analysis.

```{r}
# Convert special strings to factors (if not already) and include special values as levels
special_values <- c("[Not Available]", "[Not Applicable]", "[Discrepancy]", "[Not Evaluated]", "[Unknown]")

# Apply the transformation and save it as 'TCGA_clean'
TCGA_clean <- TCGA %>%
  mutate(across(where(is.character), factor)) %>%
  mutate(across(where(is.factor), ~fct_expand(.x, special_values)))

```

### Dropping values

Columns like 'residual_tumor' has a very high percentage of missing values, actually all rows in this case are 'NA' - not useful for analysis. If 80 % or more rows in a column are missing we drop them:

```{r}
# Define a threshold for dropping columns (80% missing values)
threshold <- 0.8 * nrow(TCGA)

# Drop columns with missing values above the threshold
cols_to_drop <- missing_values_detailed %>%
  filter(missing_count > threshold) %>%
  pull(attribute)

TCGA_clean <- TCGA %>%
  select(-all_of(cols_to_drop))
```

### Imputing values

Replacing missing/NULL data with substituted values. When data is missing in a dataset, it can hinder analysis. Therefore, imputations is performed as a preprocessing step to allow for further analysis.

```{r}
# Impute missing values for numerical columns with median
TCGA_clean <- TCGA_clean %>%
  mutate(across(where(is.numeric), ~if_else(is.na(.), median(., na.rm = TRUE), .)))

# Impute missing values for categorical columns with mode
mode <- function(x) {
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}

TCGA_clean <- TCGA_clean %>%
  mutate(across(where(is.character), ~if_else(is.na(.), mode(.), .)))



# Check if there are still any missing values
sum(is.na(TCGA_clean))
sum(is.na(TCGA))
```

After cleaning missing values, what is still an issue:

```{r}
# Get a summary of missing values
missing_summary <- TCGA_clean %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(cols = everything(), names_to = "column", values_to = "n_missing")

# Check columns with remaining missing values
missing_summary %>%
  filter(n_missing > 0)

```

## Renaming columns

```{r}
# Renaming column names for better understanding
TCGA_clean <- TCGA_clean %>%
  rename(
    sampleID = bcr_patient_barcode,
    cancer_type = type,
    age_diagnosis = age_at_initial_pathologic_diagnosis,
    pathologic_stage = ajcc_pathologic_tumor_stage
  )
```

## Saving the cleaned data

```{r}
# Save the clean data
write_tsv(TCGA_clean, "~/project29_git/data/02_dat_clean.tsv")
```
